File-based localization of numerical perturbations in data analysis pipelines

### 2020-08-28 paper.tex
update readme and bib files

Numerical evaluations are traditionally performed using techniques such as
interval arithmetics~\cite{hickey2001interval} that require complete code re-writes and are
therefore barely applicable to complex pipelines. Recently, Monte-Carlo
Arithmetic~\cite{Parker1997-qq, Denis2016-wo} provided a practical way to
evaluate the uncertainty of numerical results without the need to rewrite
the application in a different paradigm. By perturbating floating-point
computations, it introduces a controllable amount of noise in the
pipelines, effectively sampling results from a random distribution. While
this technique is very appealing, it suffers from two main issues that make
it impractical at the scale of a complete pipeline. First, it requires that
all software components be recompiled for MCA instrumentation, which is not
always feasible. Second, it multiplies the execution time by a factor of 10
to 100, which is impractical when executions already take a few hours to
complete.

@@

\subsection{Key findings}
Linear and non-linear registration with FSL were found to
frequently lead to differences between results obtained with different
operating systems. This does not come as a surprise given the instabilities
associated with these processes. It also corroborates our previous findings
in~\cite{Glatard2015}, where fMRI pre-processing with FSL was found to vary across operating systems
starting from the motion
correction step, a step that uses FSL's \flirt tool internally. It
would be relevant to investigate if the observed instability of
registration processes generalizes to other toolkits, or if it remains specific
to FSL. In view of the effect of small data perturbations in a variety of
toolboxes and processes, such as cortical surface extraction using
FreeSurfer and CIVET~\cite{Lewis2017-ll} or connectome estimation using
Dipy~\cite{kiar2019comparing}, it is probable that this observation
generalizes widely across toolboxes and requires a deeper investigation of
the stability of linear and non-linear registration.

@@

Our results illustrate the type of variability that can be introduced in
 neuroimaging results due to operating system updates. The
 numerical noise introduced by operating system updates is realistic, as
 such updates are likely to occur throughout the time span of a
 neuroscience study, but it is also uncontrolled, as it originates in
 updates of low-level libraries by third-party developers. A possible
 method to study this problem more comprehensively would be to introduce
 controlled numerical perturbations in pipelines, which could be done by
 introducing noise either in the data, or in floating-point computations
 through Monte-Carlo Arithmetic~\cite{Parker1997-qq}. The work
 in~\cite{kiar2019comparing} discusses and compares these two techniques.

@@

\revised{
\toolname only tests pipeline reproducibility in the scope of a particular dataset. However, it is very plausible 
for pipeline processes to exhibit different reproducibility behaviors when executed on different datasets. 
Therefore, only the lack of reproducibility of a pipeline process could be guaranteed from an analysis with \toolname, 
since proving reproducibility would require testing the pipeline on all possible datasets, in all possible environments, 
which is not feasible. Two elements could be considered in future work to address this issue. 
First, similar to conventional software testing, a code coverage metric could be developed to assess 
the fraction of the pipeline code involved in the tested dataset and parameters. This would quantify 
the representativity of the dataset and pipeline parameters used in the evaluation. Second, statistical 
risk models could be used to estimate the probability for a process to be reproducible, given a set of 
observations with no numerical differences. For instance, models similar to the work in~\cite{beauzamy2004methodes} 
could be leveraged for this purpose.}

>>>>

Numerical evaluations are traditionally performed using techniques such as
interval arithmetics~\cite{hickey2001interval} that require complete code re-writes and are
therefore barely applicable to complex pipelines. Recently, Monte-Carlo
Arithmetic~\cite{Parker1997-qq, denis2015verificarlo} provided a practical way to
evaluate the uncertainty of numerical results without the need to rewrite
the application in a different paradigm. By perturbating floating-point
computations, it introduces a controllable amount of noise in the
pipelines, effectively sampling results from a random distribution. While
this technique is very appealing, it suffers from two main issues that make
it impractical at the scale of a complete pipeline. First, it requires that
all software components be recompiled for MCA instrumentation, which is not
always feasible. Second, it multiplies the execution time by a factor of 10
to 100, which is impractical when executions already take a few hours to
complete.

@@

\subsection{Key findings}
Linear and non-linear registration with FSL were found to
frequently lead to differences between results obtained with different
operating systems. This does not come as a surprise given the instabilities
associated with these processes. It also corroborates our previous findings
in~\cite{Glatard2015}, where fMRI pre-processing with FSL was found to vary across operating systems
starting from the motion
correction step, a step that uses FSL's \flirt tool internally. It
would be relevant to investigate if the observed instability of
registration processes generalizes to other toolkits, or if it remains specific
to FSL. In view of the effect of small data perturbations in a variety of
toolboxes and processes, such as cortical surface extraction using
FreeSurfer and CIVET~\cite{Lewis2017-ll} or connectome estimation using
Dipy~\cite{kiar2020comparing}, it is probable that this observation
generalizes widely across toolboxes and requires a deeper investigation of
the stability of linear and non-linear registration.

@@

Our results illustrate the type of variability that can be introduced in
 neuroimaging results due to operating system updates. The
 numerical noise introduced by operating system updates is realistic, as
 such updates are likely to occur throughout the time span of a
 neuroscience study, but it is also uncontrolled, as it originates in
 updates of low-level libraries by third-party developers. A possible
 method to study this problem more comprehensively would be to introduce
 controlled numerical perturbations in pipelines, which could be done by
 introducing noise either in the data, or in floating-point computations
 through Monte-Carlo Arithmetic~\cite{Parker1997-qq}. The work
 in~\cite{kiar2020comparing} discusses and compares these two techniques.

@@

\revised{
\toolname only tests pipeline reproducibility in the scope of a particular dataset. However, it is very plausible 
for pipeline processes to exhibit different reproducibility behaviors when executed on different datasets. 
Therefore, only the lack of reproducibility of a pipeline process could be guaranteed from an analysis with \toolname, 
since proving reproducibility would require testing the pipeline on all possible datasets, in all possible environments, 
which is not feasible. Two elements could be considered in future work to address this issue. 
First, similar to conventional software testing, a code coverage metric could be developed to assess 
the fraction of the pipeline code involved in the tested dataset and parameters. This would quantify 
the representativity of the dataset and pipeline parameters used in the evaluation. Second, statistical 
risk models could be used to estimate the probability for a process to be reproducible, given a set of 
observations with no numerical differences. For instance, models described in~\cite{beauzamy2004methodes} 
could be leveraged for this purpose.}
